---
title: "Creating an R Package"
author: "Kelsey Grinde"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{creating-a-package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE
)
```

```{r setup}
library(toyr)
```

# References

Information for this report is pulled from 

- this [book](http://r-pkgs.had.co.nz/intro.html) (and an [updated version](https://r-pkgs.org/intro.html) of that book) by Hadley Wickham
- this [blog post](https://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/) by Hilary Parker
- this GitHub [help page](https://docs.github.com/en/desktop/contributing-to-projects/adding-an-existing-project-to-github-using-github-desktop)

# Setup

Before you get started, you may need to install and/or update some applications and R packages. 

## Update R and RStudio

You'll need recent versions of R (at least 3.5.0) and RStudio (at least 1.2.5042). 

To check what versions you're currently running, you'll first need to install the `rstudioapi` package (if you haven't already). Run this code chunk to install:

```{r install-rstudioapi, eval = FALSE}
install.packages("rstudioapi")
```

Then you can use these functions to check what versions of R and RStudio you have:

```{r check-versions, eval = FALSE}
getRversion()  # at least 3.5.0
rstudioapi::versionInfo()$version # at least 1.2.5042
```

If either of these needs updating, do the following:

- Update R: [download](https://www.r-project.org/) the new version 
- Update RStudio: click `Help` > `Check for Updates`

## Install R Packages

There are a number of R packages that make creating a *new* R package much easier. 

You probably already have two of these packages: `knitr` and `rmarkdown`. If not, run the code chunk below:

```{r install-packages-part1, eval = FALSE}
install.packages(c("knitr", "rmarkdown"))
```

We also need three other packages which you more likely have not yet installed. Run this code chunk to install: 

```{r install-packages, eval = FALSE}
install.packages(c("devtools", "roxygen2", "testthat"))
```

## Install Rtools or XCode 

Finally, you'll need some additional tools that can be installed as follows:

- Windows: download and install [Rtools](https://cran.r-project.org/bin/windows/Rtools/)
- Mac: make sure you have XCode (free in the App Store) or the [Command Line Tools for XCode](http://developer.apple.com/downloads)

## Confirm Setup Was Successful

Once you've done all of the above, you can check that everything went smoothly by running the following code:

```{r check-setup, eval = FALSE}
library(devtools)
has_devel()
```

If all is okay, this should return `TRUE` (and may also print the message `Your system is ready to build packages!`). 


# Creating a Minimially Viable Package

## Getting Started

Getting a package started is very easy, thanks to RStudio! Follow these steps:

0. Come up with a name for your package: can consist of letters and numbers, and must start with a letter; other suggestions [here](https://r-pkgs.org/workflows101.html#naming)
1. Click `File` > `New Project`
2. Choose `New Directory`
3. Then `R Package`
4. Type in the name of your package under `Package name:`
5. Click `Create Project`


This will create a new package containing six components:

- A `DESCRIPTION` file, which stores important metadata about your package (you'll edit this eventually)
- A `NAMESPACE` file, which helps your package interact nicely with other packages (you will not edit this directly) 
- An `R/` directory, which is where you'll store all R code for your package (you'll add files to this)
- A `man/` directory, which stores important documentation for each of the objects in your package (like the `NAMESPACE`, you won't edit this directly but it will be updated via the process described below)
- A `pkgname.Rproj` file, which makes your package easy to use with RStudio (you will not edit this)
- A `.Rbuildignore` file, which tells R which directories/files are in this directory that you do not want to actually be included in the package bundle (you may edit this eventually)


## Adding a Function

Suppose you want to add the following function to your package:

```{r, eval = FALSE}
toy_function <- function() {
  print("This is a function")
}
```

To add this to your package, create a new file in the `R/` directory called `toy_function.R`:

1. Click `File` > `New File` > `R Script` 
2. Copy the function above and past it into the new file
3. Save the file (call it `toy_function.R` and make sure it's saved in the `R/` directory)

*Note: you can also accomplish Steps 1 and 3 by running the code chunk above to define `toy_function` and then running `use_r("toy_function")`. This will create and open `R/toy_function.R` and then you just need to complete Step 2 and save.*


## Testing the Function

To test whether the new `toy_function` works, call `load_all()` to make the function available for interactive use:

```{r, eval = FALSE}
load_all()
```

then call `toy_function()` to see how it works:

```{r}
toy_function()
```

Note that `load_all()` simulates the process of building, installing, and loading the `toyr` package, rather than just adding the functions in the `toyr` package to your Global Environment so you can work with them (which would be another option).

## Documenting the Function

Thanks to the `roxygen2` package, it's fairly easy to add documentation to your package for your new function. You'll do this by adding comments to the beginning of each function; those comments will later be compiled into correctly formatted package documentation.

The general workflow for adding documentation is as follows:

1. Add roxygen comments to `.R` files
2. Run `devtools::document()` to convert those comments to `.Rd` files in the `man/` directory
3. Preview documentation using `?functionname`
4. Repeat until you're satisfied with how the documentation looks

You'll use roxygen comments to provide the following information for the function:

- what it is (TITLE)
- what it does (DESCRIPTION)
- what arguments it takes (@param)
- what outputs it returns (@return)
- how it works (@examples)
- what other functions or resources it's related to (@seealso)
- whether it's visible to the user (@export)

Roxygen comments always start with the characters `#'`.  Here's the structure of a roxygen block for a typical function:

```{r, eval = FALSE}
#' A short (one line) title.
#' 
#' Here is where you should briefly describe (less than one
#' paragraph) what the function is supposed to do.
#' 
#' @param arg1 A description of the type of the first parameter
#'   (e.g., string, numeric vector) and what it does
#' @param arg2 A description of the second parameter. If it has
#'   has a default value, you should also mention that here.
#' 
#' @return A description of the output from the function.
#' 
#' @examples
#' function(arg1 = value1, arg2 = value2)
#' 
#' @seealso This website \url{http://useful-url.com}, this
#'   this function in our package \code{\link{functionname}}, 
#'   and this one in another package \code{\link[pkgname]{functionname}}
#'
#' @export
```

For more details, read [this](http://r-pkgs.had.co.nz/man.html).

To add documentation to your `toy_function` function:

1. Add roxygen comments to your `toy_function.R` file so that it now looks like this:  

```{r, eval = FALSE}
#' A toy function. 
#' 
#' This function prints out a simple, not very useful, message.
#' 
#' @return None. The function simply prints out a message.
#' 
#' @examples
#' toy_function()
#' 
#' @seealso The general \code{\link[base]{print}} function.
#' 
#' @export
toy_function <- function() {
  print("This is a function")
}
```

2. Process this documentation by running:
```{r, eval = FALSE}
devtools::document()
```
3. Check that this created a new `toy_function.Rd` file in the `man/` directory. It should also add "export(toy_function)" to the `NAMESPACE` file (if this didn't happen, delete the `NAMESPACE` file and try Step 2 again.)
4. Preview the documentation by running `?toy_function`. This should open a help page populated with the (nicely formatted) information from your roxygen comments.
5. Repeat Steps 1-4, as needed.


## Adding Comments for the Package

You can also use roxygen comments to create a help page for your package as a whole, which will be accessible using `package?pkgname`. Use this to describe the most important components of your package (but you'll probably want to supplement with vignettes, as well). 

First, you'll need to create new file in the `R/` directory called `pkgname.R`.

Then, fill out this file with roxygen comments, attached to a `NULL` object, according to this general structure: 

```{r, eval = FALSE}
#' pkgname: A title for the package.
#'
#' Here is where you should describe what the 
#' package is supposed to do. For example, it is 
#' useful to describe what categories of functions 
#' the package provides.
#' 
#' @docType package
#' @name pkgname
NULL
```


For example, here's what you might put in your `toyr.R` file:

```{r, eval = FALSE}
#' toyr: A toy package.
#'
#' This package contains one simple, not very 
#' useful function. It was created for demonstration
#' purposes. 
#' 
#' @docType package
#' @name toyr
NULL
```

Create and update this `toyr.R` file, and then run 

```{r, eval = FALSE}
devtools::document()
````

This should produce a new file `man/toyr.Rd`. 


## Checking the Package

While you're working on the package, and before wrapping things up, it's a good idea to run `check()` to make sure everything looks okay:

```{r, eval = FALSE}
devtools::check()
```

## Installing the Package

Now, we're finally ready to install our R package.

*Note: when we created the R package, RStudio autopopulated the `R/` and `man/` directories with two files: `hello.R` and `hello.Rd`, respectively. You don't need these files for your package, so go ahead and delete them now.*

Install the package by running:

```{r, eval = FALSE}
devtools::install()
```

Try typing `?toy_function` to see if the standard help pages pop up.


# Putting the Package on GitHub

First, we need to make the package a git repository:

1. Click `Tools` > `Version Control` > `Project Setup`
2. Set the version control system from "None" to "Git"
3. Click "Yes" (we want to create a new git repository).
4. Click "Yes" to restart RStudio.

You should now see a `Git` tab in the `Environment`/`History`/etc. panel.  Click the `Commit` button, select the files you want to add to your first commit (probably all of them), type in a short commit message, and click `Commit`.

After this initial commit, you're ready to put this package on GitHub:

1. Open GitHub Desktop
2. Click `File` > `Add local repository`
3. Click `Choose`
4. Navigate to the location of this R package
5. Click `Add repository`
6. Click the `Publish repository` button on the toolbar
7. Make sure the name of the repository is the same as the name of your package
8. Uncheck `Keep this code private` if you want the repository to be public
9. Choose the organization where you want to publish the repository (or None if you want to publish to your personal account)
10. Click `Publish repository`

You should now be able to see your repository on GitHub. Take a look to make sure.

## Installing a Package from GitHub

*Note: your repository needs to be public in order for this to work*

Run `install_github("usrname/pkgname")`. For example:

```{r, eval = FALSE}
library(devtools)
install_github("GrindeLab/toyr")
```

Note that when you install a package from GitHub using `install_github`, it by default will not build vignettes. To force this to happen, run:

```{r, eval = FALSE}
devtools::install_github('GrindeLab/toyr', build_vignettes = TRUE)
```

# Updating the Package

## Editing the Package Description

The `DESCRIPTION` file contains important metadata about your package. At minimimum, you should make the following edits:

1. Give your package a `Title`. Follow these recommendations: 
    - should be a short, one line description of your package
    - $\le$ 65 characters
    - use title case
    - don't use any markup
    - don't end with a period
    - e.g., "Toy Package for Making R Packages"
2. Make yourself the author. Follow these recommendations: 
    - use the `Authors@R` field instead of `Author` and `Maintainer`
    - use the R code `person("First", "Last", email = "email@email.com", role = c("aut", "cre"))` to list yourself as both the author (aut) and maintainer (cre)
    - you can also list multiple authors; more info [here](https://r-pkgs.org/description.html#author)
3. Write a description of your package in the `Description` field.  
    - this should be more detailed than the title
    - you can use multiple sentences, but no more than one paragraph
    - you should split text across multiple lines, with no more than 80 characters per line
    - indent new lines with 4 spaces
4. Add a license to indicate who can use your package (and how they can use it).  
    - the [MIT](https://tldrlegal.com/license/mit-license) is often used; it lets people use and distribute your code freely as long as the license is always distributed with the code
    - to use the MIT license run `use_mit_license("Your Name")`; this will add "MIT + file LICENSE" to the `License` field of the `DESCRIPTION` file; it also creates the `LICENSE` and `LICENSE.md` files with more details about the license, and adds `LICENSE.md` to the `.Rbuildignore` file
    - more details, including other license options, [here](https://r-pkgs.org/whole-game.html#use_mit_license) and [here]()

Once you've made these updates, check your package:

```{r, eval = FALSE}
devtools::check()
```

You should no longer see a `WARNING` message about having a non-standard license.

Assuming everything looks okay, install the package:

```{r, eval = FALSE}
devtools::install()
```


## Creating Vignettes

(under construction)

To create your first vignette, run: 

```{r, eval = FALSE}
usethis::use_vignette("my-vignette")
```

This will:

- Create a new directory called `vignettes/` 
- Update the `DESCRIPTION` file 
- Created a draft vignette `vignettes/my-vignette.Rmd`


To create future vignettes, add a new `.Rmd` file to the `vignettes/` directory. Use the following structure for the YAML header:

```
---
title: "Vignette Title"
author: "Your Name"
date: "Date of Last Update"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```

Edit this `.Rmd` file and `Knit` as you would with any other RMarkdown file. 

After adding your content, run the following to build the vignette.

```{r, eval = FALSE}
devtools::document()
devtools::build()
```

Then, install the package by running:

```{r, eval = FALSE}
devtools::install()
```
Check for vignettes by running:

```{r, eval = FALSE}
browseVignettes('toyr')
```

(If you're getting a `no vignettes found` error message here, you may need to `Install` > `Install from: Package Archive File`. Then try running `browseVignettes` again.)

## Adding a README

To initialize a basic `README.Rmd` file, run `use_readme_rmd()`:

```{r, eval = FALSE}
usethis::use_readme_rmd()
```

Open the newly created `README.Rmd`. Update the following:

- describe the purpose of your package (see the pkgname section) (feel free to copy from your `DESCRIPTION` file)
- update the installation information (e.g., delete the info about CRAN if your package is not on there yet)
- show some examples of how to use your package

Once you've updated this RMarkdown file to your satisfaction (and saved your changes), you'll need to render it to make `README.md`:

```{r, eval = FALSE}
rmarkdown::render("README.Rmd")
```

I'd recommend adding "README.html" to the list of files that get ignored when you build your package (otherwise you may get a note about it from the `check()` function). Here's how to add tell R to ignore this file:  

```{r, eval = FALSE}
usethis::use_build_ignore("README.html")
```

Next, you should commit and push to GitHub. 

## Updating .Rbuildignore

You may have noticed in some of the steps above that `.Rbuildignore` is sometimes automatically updated by `devtools` functions.

Otherwise, there may be files that you've created that you don't actually want included in your R package. For example, I have two files, `originstructions.Rmd` and `originstructions.html`, that I had originally used to document the steps  for creating a package; however, I have since moved that information to this vignette, so I do not want these original files to be included in the R package. Here's how I added them to `.Rbuildignore`:

```{r, echo = FALSE, eval = FALSE}
usethis::use_build_ignore("originstructions.html")
usethis::use_build_ignore("originstructions.Rmd")
```

```{r, eval = FALSE}
# update package and check
devtools::document()
devtools::install()
devtools::check()
```

## Adding Data

(under construction)


